An Efficient Parameter Space Search as an Alternative to Markov Chain Monte Carlo

### 2014-07-20 aps4p5/drafts/draft_140719.tex
edited the draft

Bryan {\it et al.} (2007) show that this algorithm is robust against multi-modal
likelihood functions, using it to identify a disjoint region of high likelihood in
cosmological parameter space as constrained by the 1-year WMAP CMB anisotropy power
spectrum.  This is the principal advantage to selecting points for evaluation according
to the statistic (\ref{eq:sstat}).  Unfortunately, it is possible that the focus on exploring
unknown regions of parameter space might slow down convergence in the case of uni-modal
likelihood functions.  Therefore, we propose the following modifications to the algorithm
to ensure that it adequately characterizes known regions of high likelihood while
simultaneously searching for new regions to explore.

\subsubsection{Characterization of known high likelihood regions}
\label{sec:focus}

As the algorithm searches, it will keep track of a list of all of the discovered points
which are believed to be local minima of $\chi^2$ (how it finds these will be described below
in Section \ref{sec:simplex}).  After each cycle of the main algorithm (steps 1A-3A above),
our version of APS attempts ``focused searches'' about these local minima.

For each local minimum $\vec{M}$, APS performs the following:
\begin{itemize}
\item (1B) If fewer than $N_p$ focused searches have been proposed around this minimum, randomly
choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
evaluate $\chi^2$ at that point.  
\\
\item(2B) If the point does not satisfy $\chi^2=\chi^2_\text{lim}$ (which is highly probable), 
use bisection
along the line connecting the proposed point to $\vec{M}$ to find a point which does
satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line with
$\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
with $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered this way with
$\chi^2=\chi^2_\text{lim}$ are stored as ``boundary points'' about $\vec{M}$.  Each local
minimum will have its own list of boundary points.
\\
\item(3B) If $N_p$ boundary points have been found for $\vec{M}$, skip steps (1B) and (2B).
Instead, examine the boundary points associated with $\vec{M}$ and identify the points
corresponding to both the maximum and minimum values of each parameter $\theta_i$.  These
points will be referred to as $\vec{B}^{(j)}$.  There will be $2\times N_p$ of them, one for
the maximum and one for the minimum in each of our $N_p$ parameters.  
\\
\item(4B) For each point
$\vec{B}^{(j)}$, propose some number $N_v$ of new points (we have chosen
$N_v=20$ for the sake of speed) $\vec{d}^{(k)} = \vec{B}^{(j)}+\vec{v}$ where $\vec{v}$ is a
random, small vector in parameter space constrained to be perpendicular to
$\vec{B}^{(j)}-\vec{M}$.  There will now be $2\times N_p \times N_v$ points $\vec{d}^{(k)}$
($N_v$ such points for each of the $2\times N_p$ extremal boundary points $\vec{B}^{(j)}$).
\\
\item(5B) At each point $\vec{d}^{(k)}$, evaluate equation (\ref{eq:sstat}).  Evaluate $\chi^2$
at the point $\vec{d}^{(k)}$ which maximizes $S$. If
$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
\end{itemize}

@@

\subsubsection{Extending the boundary}
\label{sec:unitSphere}

The modification suggested in section \ref{sec:focus} is designed to explore 
$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
perpendicularly to its ``radii'' (the vector $\vec{B}^{(j)}-\vec{M}$ would be a radius of
$\chi^2=\chi^2_\text{lim}$ described a sphere).

something about APS wide helping to discover new corners of the surface as well

@@

>>>>

\subsection{Modifications to the APS algorithm in the present work}

Bryan {\it et al.} (2007) show that the algorithm above 
is robust against multi-modal
likelihood functions, using it to identify a disjoint region of high likelihood in
cosmological parameter space as constrained by the 1-year WMAP CMB anisotropy power
spectrum.  This is the principal advantage to selecting points for evaluation according
to the statistic (\ref{eq:sstat}).  Unfortunately, it is possible that the focus on exploring
unknown regions of parameter space might slow down convergence in the case of uni-modal
likelihood functions.  Therefore, we propose the following modifications to the algorithm
to ensure that it adequately characterizes known regions of high likelihood while
simultaneously searching for new regions to explore.

We are greatly indebted to Eric Linder of the Lawrence Berkeley National
Laboratory who first suggested to us the modifications described below
in sections \ref{sec:focus} and \ref{sec:simplex}.

\subsubsection{Characterization of known low-$\chi^2$ regions}
\label{sec:focus}

As it searches, APS will keep track of a list of all of the discovered points
which are believed to be local minima of $\chi^2$ (how it finds these will be described below
in Section \ref{sec:simplex}).  These points will be stored in the set
$\{\Gamma\}$.  After each cycle of the main algorithm (steps 1A-3A above),
our version of APS attempts ``focused searches'' about these local minima.

For each local minimum $\vec{M}$ in $\{\Gamma\}$, APS performs the following:
\begin{itemize}
\item (1B) If fewer than $N_p$ focused searches have been proposed around this minimum, randomly
choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
evaluate $\chi^2$ at that point.  
\\
\item(2B) If the point does not satisfy $\chi^2=\chi^2_\text{lim}$ (which is highly probable), 
use bisection
along the line connecting the proposed point to $\vec{M}$ to find a point which does
satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line with
$\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
with $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered this way with
$\chi^2=\chi^2_\text{lim}$ are stored as ``boundary points'' about $\vec{M}$.  Each local
minimum will have its own list of boundary points.
\\
\item(3B) If $N_p$ boundary points have been found for $\vec{M}$, skip steps (1B) and (2B).
Instead, examine the boundary points associated with $\vec{M}$ and identify the points
corresponding to both the maximum and minimum values of each parameter $\theta_i$.  These
points will be referred to as $\vec{B}^{(j)}$.  There will be $2\times N_p$ of them, one for
the maximum and one for the minimum in each of our $N_p$ parameters.  
\\
\item(4B) For each point
$\vec{B}^{(j)}$, propose some number $N_v$ of new points (we have chosen
$N_v=20$ for the sake of speed) $\vec{d}^{(k)} = \vec{B}^{(j)}+\vec{v}$ where $\vec{v}$ is a
random, small vector in parameter space constrained to be perpendicular to
$\vec{B}^{(j)}-\vec{M}$.  There will now be $2\times N_p \times N_v$ points $\vec{d}^{(k)}$
($N_v$ such points for each of the $2\times N_p$ extremal boundary points $\vec{B}^{(j)}$).
\\
\item(5B) At each point $\vec{d}^{(k)}$, evaluate equation (\ref{eq:sstat}).  Evaluate $\chi^2$
at the point $\vec{d}^{(k)}$ which maximizes $S$. If
$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
\end{itemize}

@@

\subsubsection{Extension of the $\chi^2=\chi^2_\text{lim}$ contour}
\label{sec:unitSphere}

The modification suggested in section \ref{sec:focus} is designed to explore 
$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
perpendicularly to its ``radii'' (the vector $\vec{B}^{(j)}-\vec{M}$ would be a radius of
$\chi^2=\chi^2_\text{lim}$ described a sphere).  It is, however, also possible
that the $S$-maximization in steps (1A-3A) could also learn something about the
shape of the $\chi^2=\chi^2_\text{lim}$ contour, even if it does not place
points directly on that contour.  To explore that possibility, we add the
following steps to the end of the sequence (1A-3A).

\begin{itemize}
\item(1C) For each point $\vec{M}$ in $\{\Gamma\}$, store, 
in addition to the boundary
points associated with $\vec{M}$, the projection of those boundary points
onto a unit sphere in parameter space, i.e. for each boundary point, draw a line
connecting that boundary point with $\vec{M}$ and keep track of
where a unit sphere centered on $\vec{M}$ intersects that line.  Call this set
of projectsion $\vec{s}^{(i)}$.
\\
\item(2C) Call the point for which $\chi^2$ has been evaluated in step (3A)
$\vec{P}$., find the
point $\vec{M}$ in $\{\Gamma\}$ that is nearest to $\vec{P}$ in parameter space.
\\
\item(3C) Find the
projection of $\vec{P}$ onto the unit sphere surrounding $\vec{M}$.  
Call this projected point $\vec{u}$.  Find the point among
$\vec{s}^{(i)}$ that is nearest to $\vec{u}$.  Record the distance between them.
Keep track of this distance in the set $\{D\}$ ($\{D\}$ will be a running list
of the $\vec{u}-\vec{s}^{(i)}$ distance for all points $\vec{P}$ found by steps
1A-3A).
\\
\item(4C) If the distance between $\vec{u}$ and its nearest neighbor in
$\vec{s}^{(i)}$ is greater than the 2/3 quantile of $\{D\}$,
then perform bisection between $\vec{M}$ and the point evaluated in step (3A).
\end{itemize}

This way, points on the $\chi^2=\chi^2_\text{lim}$ contour can be found whose
directions away from $\vec{M}$ in parameter space have not yet been explored.
Because we expect the unit sphere surrounding $\vec{M}$ to become more densely
populated as the algorithm runs, we periodically delete the list $\{D\}$ so that
new directions away from $\vec{M}$ can still be explored.

@@

The final modification to the original APS algorithm which we propose in this
work is to utilize function minimization to identify local minima in $\chi^2$.
One of the principal drawbacks of the algorithm presented in steps (1A-3A) is
that it makes no account of ``near misses.''  If step (3A) finds that $\chi^2$
is lower than average but still greater than $\chi^2_\text{lim}$, the original
APS algorithm does nothing with that information.  We have tried to improve upon
that with our incorporation of bisection in sections \ref{sec:focus} and
\ref{sec:unitSphere} above.  Our final innovation is to allow APS to use points
with 
\begin{equation}
\label{eqn:simplexCriterion}
\chi^2_\text{lim}\le\chi^2\le\chi^2_\text{sig}
\end{equation}
to seed a simplex function minimizer as developed by Nelder and Mead (1965).

We set
$\chi_\text{sig}=\chi^2_\text{min}+G\times(\chi^2_{1/10}-\chi^2_\text{min})$
where $G$ is a user-set parameter, $\chi^2_\text{min}$ is the minimum discovered
$\chi^2$ value and $\chi^2_{1/10}$ is the 1/10 quantile of the $\chi^2$ values
discovered by steps (1A-3A) (not those discovered by any of the modifications
discussed in sections \ref{sec:focus}, \ref{sec:unitSphere} or
\ref{sec:simplex}).

Because the Nelder-Mead simplex minimizer can be very time intensive, we balance our
algorithm by demanding that the number of $\chi^2$ evaluations devoted to all
other variants of search (i.e. steps (1A-3A) and (1B-5B)) equal the number of
$\chi^2$ evaluations devoted to the simplex minimizer before a new simplex
minimization can occur, and if no new points are known that satisfy the
criterion in equation \ref{eqn:simplexCriterion} above, no simplex minimization
will be undertaken, regardless.

If the simplex minimizer converges to a point such that
$\chi^2<\chi^2_\text{lim}$, our APS determines whether or not it
has found a new local minimum to be added to $\{\Gamma\}$ by comparing the end
point $\vec{\gamma}$ of the minimization to each of the points already in
$\{\Gamma\}$.  For each $\vec{M}$ in $\{\Gamma\}$, APS calculates the mid-point
between $\vec{\gamma}$ and $\vec{M}$.  If, for all $\vec{M}$, 
$\chi^2$ at that mid-point is greater
than $\chi^2_\text{lim}$, then it is deemed that $\vec{\gamma}$ is a new local
minimum and is added to $\{\Gamma\}$.  If not, $\vec{\gamma}$ is assumed to be a
part of the one of the previously identified loci of low $\chi^2$ and it is not
added to $\{\Gamma\}$.
